---
title: "Mining the National Diet and Nutrition Survey Rolling Programme (NDNS RP) data"
subtitle: "When, what, where you eat."
author: "王　超辰 | Chaochen Wang"
date: "2019-09-17 18:00~19:30 @AMU・疫学懇話会"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      countdown: 60000
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  # text_font_google   = google_font("M PLUS Rounded 1c", "300", "300i"),
  text_font_google = google_font("Noto Sans JP", "300", "300i"), 
  code_font_google   = google_font("Droid Mono"),
  text_bold_color = "#B03A2E",
  link_color = "#2c7fb8",
  title_slide_background_image = "pic/",
  inverse_text_shadow = TRUE
)
# write_xaringan_theme(text_bold_color = "#FF4333")
```

class: middle

# Outline of today's talk 

--
### Correspondence analyses:


- Relationship between **food consumed** and **eating location** for UK adolescents


- Relationship between **food consumed** and **eating time** for UK adults according to their diabetes status 

--

### Multilevel Latent Class Analysis:

- Relationship between **carbohydrate consumption** and **eating time** for UK adults. 

---
class: middle

# Some intuition about correspondence analysis  (CA)

--
- CA is a method for investigating the relationship in a **two-dimensional contingency table**.

- For example, in the following table gives the frequency of consumption of three healthy food groups at each location in the NDNS RP data among teenagers (age: 11~19 years):

```{r eval = F}
library(knitr)
library(kableExtra)


dt <- 
```



```
                   Locat_type
mfgLab               Others   Home  School_work  Total
  Fruit                 455   2595          839   3889
  Salad and raw veg     662   2556          607   3825
  Veg not raw           698   6122          376   7196
  Total                1815  11273         1822  14910
```

