---
title: "Mining the National Diet and Nutrition Survey Rolling Programme (NDNS RP) data"
subtitle: "Where, when, what you eat."
author: "王　超辰 | Chaochen Wang"
date: "2019-09-17 18:00~19:30 @AMU・疫学懇話会"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      countdown: 60000
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  # text_font_google   = google_font("M PLUS Rounded 1c", "300", "300i"),
  text_font_google = google_font("Noto Sans JP", "300", "300i"), 
  code_font_google   = google_font("Droid Mono"),
  text_bold_color = "#B03A2E",
  link_color = "#2c7fb8",
  title_slide_background_image = "pic/",
  inverse_text_shadow = TRUE
)
# write_xaringan_theme(text_bold_color = "#FF4333")
```

class: middle

# Outline of today's talk 

--
### Correspondence analyses:


- Relationship between **food consumed** and **eating location** for UK adolescents


- Relationship between **food consumed** and **eating time** for UK adults according to their diabetes status 

--

### Multilevel Latent Class Analysis:

- Relationship between **carbohydrate consumption** and **eating time** for UK adults. 

---
class: middle

# Some intuition about correspondence analysis  (CA)

--
- CA is a method for investigating the relationship in a **two-dimensional contingency table**.

- For example, the following table gives the frequency of consumption of three healthy food groups at each location in the NDNS RP data among teenagers (age: 11~19 years):

```{r eval = T, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
load("../CA-NDNSRP/H_teenFood.Rdata")

TableFoogGroup <- H_teenFood %>% 
  ungroup() %>% 
  group_by(mfgLab) %>% 
  summarise(n = n(), meanHpoint = mean(H_points, na.rm = T), mfgCalories = sum(Energykcal)) %>% 
  arrange(-mfgCalories) %>% 
  mutate(n.freq = paste0(round(100 * n/sum(n), 2), "%"))  %>% 
  mutate(cal.Prop = paste0(round(100 * mfgCalories/sum(mfgCalories), 2), "%"))  %>% 
  mutate(calprop = mfgCalories/sum(mfgCalories)) %>% 
  mutate(calcumprop = paste0(round(100 * cumsum(calprop), 3), "%")) %>% 
  select(-calprop) 


TableFoogGroup <- TableFoogGroup %>% 
  mutate(healthy     = meanHpoint < -2, 
         lesshealthy = meanHpoint > 4, 
         neutral     = (meanHpoint <= 4) & (meanHpoint >= -2))

# TableFoogGroup %>% 
#   filter(healthy) %>% 
#   select(-lesshealthy, -neutral, -calcumprop, -healthy)
healthyfg <- H_teenFood %>% 
  filter(mfgLab %in% c("Fruit", "Veg not raw", "Brown Bread"))
freqtab <- xtabs(~healthyfg$mfgLab + healthyfg$Locat_type)
freqtabmg <- addmargins(freqtab)
as.data.frame.matrix(freqtabmg) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, bold = T) 
```


---
class: middle 

## Usually, we will conduct a $\chi^2$ test. 

```{r eval = T, echo=FALSE, message=FALSE}
as.data.frame.matrix(freqtabmg) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, bold = T) 
```


$$
\chi^2 = \sum\frac{(\text{Observed} - \text{Expected})^2}{\text{Expected}} \sim \chi^2_{(m -1)\times(n-1)}
$$

--
### To look for evidence against the null hypothesis that **there is no difference across the columns or rows**.

---
class: middle
# Hypothesis of independence

The null hypothesis of no difference across the columns or rows is also called:

.pull-left[
- Hypothesis of independence 
]

.pull-right[
- Homogeneity assumption
]

```{r}
chisq.test(freqtab)
```


We conclude that **there is dependency between the columns (food groups) and the rows (locations)**. 

---
class: middle

## But what exactly does this dependency mean?


```{r eval = T, echo=FALSE, message=FALSE}
as.data.frame.matrix(freqtabmg) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))%>% 
  column_spec(1, bold = T) 
```

### Can we get **more information** from this contingency table? How?

--
### Yes! we can visualize their relationships!


---
class: middle 

## Step 1 - convert the frequency cell to proportions

```{r echo = FALSE}
Xp <- as.matrix(freqtab) / sum(as.matrix(freqtab))
Xpmg <- addmargins(Xp)
Xpmg <- signif(Xpmg, digits = 3)
row.names(Xpmg) <- c("Brown Bread", "Fruit", "Veg not raw", "c: col masses")
colnames(Xpmg) <- c("Others", "Home", "School_work", "r: row masses")

as.data.frame.matrix(Xpmg) %>%
  kable(caption = "Matrix X. Observed Cell Proportions") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, bold = T) 
```

- Row masses: $\mathbf{r} = \mathbf{X1}$ 

```{r}
r <- Xp %*% matrix(c(1, 1, 1), nrow = 3)
```

- Column masses: $\mathbf{c} = \mathbf{X}^T\mathbf{1}$

```{r}
c <- t(Xp)  %*% matrix(c(1, 1, 1), nrow = 3)
```

---
class: middle 
## Step 2 - The variation of each cell in matrix X from its expected value

$$\text{Deviation} = \mathbf{X} - \mathbf{rc}^T$$

```{r echo=FALSE}
Deviation <- Xp - r %*% t(c)
library(formattable)
Deviation <- addmargins(Deviation)
Dev <- sprintf("%.5f", Deviation)
Dev <- matrix(Dev, nrow = 4)
rownames(Dev) <- rownames(Deviation)
colnames(Dev) <- colnames(Deviation)
as.data.frame.matrix(Dev) %>%
  kable(caption = "Deviations from Expected") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, bold = T) 
```




---
class: middle 

## Step 2 - Calculate row profile, and column profile

.pull-left[

- Percentages of location for each food was eaten:

```{r echo=FALSE}
Dr <- diag(as.numeric(r))
Dc <- diag(as.numeric(c))
R <- solve(Dr) %*% Xp
C <- solve(Dc) %*% t(Xp)

Rmg <- addmargins(R)[-4,]
Rmg <- signif(Rmg, digits = 3)
rownames(Rmg) <- NULL
as.data.frame.matrix(Rmg) %>%
  kable(caption = "Matrix R. Row Profile") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
$$\mathbf{R} = \mathbf{D}^{-1}_r\mathbf{X}$$
]

.pull-right[
- Percentages of food was eaten at each location:

```{r echo = FALSE}
C <- solve(Dc) %*% t(Xp)

Cmg <- addmargins(t(C))[,-4]
Cmg <- signif(Cmg, digits = 3)
colnames(Cmg) <- c("Others", "Home", "School")
as.data.frame.matrix(Cmg) %>%
  kable(caption = "Matrix C. Column Profile") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  column_spec(1, bold = T) 
```
$$\mathbf{C} = \mathbf{D}^{-1}_r\mathbf{X}^T$$

]
