---
title: "Mining the National Diet and Nutrition Survey Rolling Programme (NDNS RP) data"
subtitle: "When, what, where you eat."
author: "王　超辰 | Chaochen Wang"
date: "2019-09-17 18:00~19:30 @AMU・疫学懇話会"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      countdown: 60000
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  # text_font_google   = google_font("M PLUS Rounded 1c", "300", "300i"),
  text_font_google = google_font("Noto Sans JP", "300", "300i"), 
  code_font_google   = google_font("Droid Mono"),
  text_bold_color = "#B03A2E",
  link_color = "#2c7fb8",
  title_slide_background_image = "pic/",
  inverse_text_shadow = TRUE
)
# write_xaringan_theme(text_bold_color = "#FF4333")
```

class: middle

# Outline of today's talk 

--
### Correspondence analyses:


- Relationship between **food consumed** and **eating location** for UK adolescents


- Relationship between **food consumed** and **eating time** for UK adults according to their diabetes status 

--

### Multilevel Latent Class Analysis:

- Relationship between **carbohydrate consumption** and **eating time** for UK adults. 

---
class: middle

# Some intuition about correspondence analysis  (CA)

--
- CA is a method for investigating the relationship in a **two-dimensional contingency table**.

- For example, in the following table gives the frequency of consumption of three healthy food groups at each location in the NDNS RP data among teenagers (age: 11~19 years):

```{r eval = T, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
load("../CA-NDNSRP/H_teenFood.Rdata")

TableFoogGroup <- H_teenFood %>% 
  ungroup() %>% 
  group_by(mfgLab) %>% 
  summarise(n = n(), meanHpoint = mean(H_points, na.rm = T), mfgCalories = sum(Energykcal)) %>% 
  arrange(-mfgCalories) %>% 
  mutate(n.freq = paste0(round(100 * n/sum(n), 2), "%"))  %>% 
  mutate(cal.Prop = paste0(round(100 * mfgCalories/sum(mfgCalories), 2), "%"))  %>% 
  mutate(calprop = mfgCalories/sum(mfgCalories)) %>% 
  mutate(calcumprop = paste0(round(100 * cumsum(calprop), 3), "%")) %>% 
  select(-calprop) 


TableFoogGroup <- TableFoogGroup %>% 
  mutate(healthy     = meanHpoint < -2, 
         lesshealthy = meanHpoint > 4, 
         neutral     = (meanHpoint <= 4) & (meanHpoint >= -2))

# TableFoogGroup %>% 
#   filter(healthy) %>% 
#   select(-lesshealthy, -neutral, -calcumprop, -healthy)
healthyfg <- H_teenFood[H_teenFood$HealthFoodGr == "Healthy", ]
freqtab <- xtabs(~healthyfg$mfgLab + healthyfg$Locat_type)
freqtab <- addmargins(freqtab)
as.data.frame.matrix(freqtab) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


---
class: middle 

## Usually, we will conduct a $\chi^2$ test. 

```{r eval = T, echo=FALSE, message=FALSE}
as.data.frame.matrix(freqtab) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


$$
\chi^2 = \sum\frac{(\text{Observed} - \text{Expected})^2}{\text{Expected}} \sim \chi^2_{(m -1)\times(n-1)}
$$

--
### To look for evidence against the null hypothesis that **there is no difference across the columns or rows**.

---
class: middle
# Hypothesis of independence

The null hypothesis of no difference across the columns or rows is also called:

- Hypothesis of independence 

OR 

- Homogeneity assumption


